A free/open-source Kazakh-Tatar machine translation system

### 2013-04-23 paper/paper.tex
submitted versions (full and anonymised)

\documentclass[11pt]{article}

@@

@@

\author{Ilnar Salimzyanov\\
  Kazan Federal University\\
  Kazan, Republic of Tatarstan\\
  Russian Federation\\
  \email{ilnar.salimzyan@gmail.com}  \And
  Jonathan North Washington\\
  Departments of Linguistics\\
  and Central Eurasian Studies\\
  Indiana University\\
  Bloomington, Indiana 47405 USA\\
  \email{jonwashi@indiana.edu}  \And
  Francis Morton Tyers\\
  Departament de Llenguatges\\
  i Sistemes Informàtics\\
  Universitat d'Alacant\\
  E-03877 Alacant\\
  \email{ftyers@dlsi.ua.es}}

@@

Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Tatar--Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.  Among these systems is a prototype Tatar--Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.

@@

Kazakh is primarily spoken in Kazakhstan, where it is the national
language, sharing official status with Russian as an official language.
Large groups of native speakers also exist in China, neighbouring
Central-Eurasian republics, and Mongolia. The total number of speakers is at least 10 million people.

@@

\begin{figure*}[htbp]
\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
%\begin{texttt}
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e> \\
    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
}}%\end{texttt}
\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

@@

%FIXME: Describe the gist of the rules, maybe with a few concrete examples.
The grammar currently has 60 rules.  The goal of these rules is to select the correct analysis when there 
are multiple morphological analyses of Kazakh forms.
%FIXME: It might be good to put some numbers about the morphological ambiguity of Kazakh

%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
The output of the morphological analyser is highly ambiguous. Both Kazakh and Tatar have a lot of affixes which join the
verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.

We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
adjective], [finite form], [substantivised verbal adjective, i.e., if the word it determines was omitted], whereas many grammars
do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.

This type of ambiguity is common between the two languages we deal with, and therefore may be passed from one language
to another and most likely will not lead to translation errors. Disambiguating it, and, prior to that, having that kind of
informaton about the syntactical role of a surface form is crucial for success e.g. of a Turkic to English system.

This is the reason for a high ambiguity rate of the transducers.

@@

For example (see figure~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
meanings corresponding to its Tatar cognate корал; however, it is also used in the 
compound ақпарат құралдары \eng{mass media} (literally, \eng{means of information}), which translates 
to Tatar as мәгълүмат чаралары (which has the same literal translation).  Hence, the Kazakh word құрал must have 
two entries in the bilingual lexicon: one that corresponds to Tatar корал and one that corresponds 
to Tatar чара.  A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in figure~\ref{fig:lrx}.

@@

Likewise, the Kazakh word топ has two translations into Tatar.  One translation is туп \eng{ball} (which can also 
be доп in Kazakh), and the other translation is төркем \eng{group}.  The bilingual dictionary also has 
the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated 
to Kazakh топ (i.e., analysed), but is never generated.

Another lexical selection rule is shown in figure~\ref{fig:lrx2}.  This rule ensures that the Kazakh word рет \eng{row, time(s)} is translated as тапкыр \eng{time(s)} when it is used with a number; otherwise it translates as рәт \eng{row}.
\begin{figure}
\begin{lstlisting}
  <rule>
    <match tags="num"/>
    <match lemma="рет" tags="n.*">
        <select lemma="тапкыр"/>
    </match>
  </rule>
\end{lstlisting}
\caption{A lexical selection rule that selects Tatar тапкыр for Kazakh рет when preceded by a number.}
\label{fig:lrx2}
\end{figure}

Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).  The lexical selection rule is provided in figure~\ref{fig:lrx3} chooses the latter Tatar translation after the word рухани \eng{soul}.

\begin{figure}
\begin{lstlisting}
  <rule>
    <match lemma="рухани" tags="adj"/>
    <match lemma="өмір" tags="n.*">
        <select lemma="тормыш"/>
    </match>
  </rule>
\end{lstlisting}
\caption{A lexical selection rule that selects Tatar тормыш for Kazakh өмір when preceded by the word рухани.}
\label{fig:lrx3}
\end{figure}

@@

@@

For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty's Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.

For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.

@@

%%
%%\begin{table}[htbp]
%%	\centering
%%	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
%%		\begin{tabular}{llrrr}
%%			\toprule
%%			Corpus          & Tokens & Coverage & stdev\\
%%			\midrule
%%			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
%%			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
%%			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
%%			\bottomrule
%%		\end{tabular}
%%	}\\
%%	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
%%		\begin{tabular}{llrrr}
%%			\toprule
%%			Corpus          & Tokens & Coverage & stdev\\
%%			\midrule
%%			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
%%			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
%%			New Testament   & 137K   & 92.54\%  & 1.31\% \\
%%			\bottomrule
%%		\end{tabular}
%%	}
%%	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
%%	\label{table:coverage}
%%\end{table}

\begin{table}
	\centering
	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
			RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
			Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
			\bottomrule
		\end{tabular}
	}\\
	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2007-2012 & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
			\bottomrule
		\end{tabular}
	}

	\caption{Naïve coverage of the Kazakh-Tatar system}
	\label{table:trimmedcoverage}
\end{table}

As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over 
the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
For the Wikipedia corpus, the coverage 
is substantially worse.  This is due to the fact that this corpus is ``dirtier'' --- 
containing more orthographical errors, \emph{wiki} code, and repetitions, as well as to the 
fact that it has contains many more proper nouns.

@@

We had two small Kazakh corpora along with their postedited translations into Tatar to measure the WER.
The first one (2,457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar, lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN

The second corpus --- a collection of articles from Radio Free Europe / Radio Liberty's Kazakh-language service --- was used solely for evaluation.  It had a similar size --- 2,862 words.
The WER results for this corpus are presented in table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.

Some of the corrections that were made as part of postediting were were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.  A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.  Aside from the irregular behaviour of йөр, which would need its irregularities implemented through special phonotactics in lexc, these issues are all shortcomings of our phonological layer, implemented in twol.

\begin{table}[htbp]
	\centering
	\begin{tabular}{lll}
		\toprule
		form & current & expected \\\midrule
		ди\tag{<v>}\tag{<tv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & дия & ди \\
		йөр\tag{<v>}\tag{<iv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & йөрә & йөри \\
		кияү\tag{<n>}\tag{<px3sp>}\tag{<nom>} & кияүе & кияве \\
		укы\tag{<v>}\tag{<iv>}\tag{<ger>}\tag{<nom>} & укыу & уку \\\bottomrule
	\end{tabular}
	\caption{Examples of some phonological problems in the Tatar transducer.}
	\label{tab:phonerrors}
\end{table}

@@

\begin{table}
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
  \midrule
  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over two corpora. The development corpus (\texttt{devel}) was used during development 
       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.}
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}

@@

\subsection{Error analysis}

@@

The majority of errors currently are due to one of the following: mistakes and gaps in the Tatar morphophonology component, lack of transfer rules to handle some Kazakh compounds, and disambiguation errors.
%\cite{TyersAlperen2010}

@@

To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production-level
coverage, but rather a prototype-level number of rules. Though the impact of this relatively low number of rules on the
quality of the translation (compare the WER results for the corpus 1 and corpus 2) looks very promising and suggests
the a high-quality translation between morphologically-rich agglutinative languages is possible.

We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
increase it to 95\% on the corpora we have, the main work will be improving the quality of the translation by adding
more rules, starting with the Constraint Grammar module.
The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.

>>>>

\documentclass[a4paper,11pt]{article}

@@

% for anonymising
% make a symlink to this file containing "anon" in the filename
% then put anything to be deleted in that version in \anonymise{}
% compile and enjoy
\usepackage{ifthen}
\usepackage{ifjobname}
\ifjobinfix{anon}{
	\newcommand{\anonymise}[1]{}
}{
	\newcommand{\anonymise}[1]{#1}
}

@@

\author{\anonymise{Ilnar Salimzyanov\\
  Kazan Federal University\\
  Kazan, Republic of Tatarstan\\
  Russian Federation\\
  \email{ilnar.salimzyan@gmail.com}  \And
  Jonathan North Washington\\
  Departments of Linguistics\\
  and Central Eurasian Studies\\
  Indiana University\\
  Bloomington, Indiana 47405 USA\\
  \email{jonwashi@indiana.edu}  \And
  Francis Morton Tyers\\
  Departament de Llenguatges\\
  i Sistemes Informàtics\\
  Universitat d'Alacant\\
  E-03877 Alacant\\
  \email{ftyers@dlsi.ua.es}}}

@@

Within the Apertium project, work on several MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Tatar--Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.  Among these systems is a prototype Tatar--Bashkir machine translation system \anonymise{which was built by the authors of this paper }\citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.

@@

Kazakh is primarily spoken in Kazakhstan, where it is the national
language, sharing official status with Russian as an official language.
Large communities of native speakers also exist in China, neighbouring
Central-Eurasian republics, and Mongolia. The total number of speakers is at least 10 million people.

@@

\begin{figure*}[bhtp]
\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
%\begin{texttt}
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e> \\
    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
}}%\end{texttt}
\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

@@

%FIXME: Describe the gist of the rules, maybe with a few concrete examples.

The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh.  The goal of the CG rules is to select the correct analysis when there are multiple analyses.  As of r43664, the ambiguity was down to 1.4 analyses per form.

One reason for the still high level of morphological ambiguity is a series of affixes in both languages which can each be analysed as some combination of verbal nouns, verbal adjectives, substantitivised verbal adjectives (i.e., verbal adjectives with a null modified noun), and even finite forms.\footnote{Despite the fact that the various suffixes in this category pattern differently and do not form a single natural class, most grammars of the languages label all them as simply ``gerunds'' or ``participles''.}

Given the similarity of Kazakh and Tatar, this sort of ambiguity may often be passed from one language to the other and not lead to many translation errors.  While disambiguating between these analyses would be crucial for e.g., a Turkic-to-English system, we have not yet put much effort into developing CG rules to deal with such ambiguity.

%This is the reason for a high ambiguity rate of the transducers.

@@

For example (see Fig.~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
meanings corresponding to its Tatar cognate корал; however, it is also used in the 
compound ақпарат құралдары \eng{mass media} (literally, \eng{means of information}), which translates 
to Tatar as мәгълүмат чаралары (which has the same literal translation).  Hence, the Kazakh word құрал must have 
two entries in the bilingual lexicon: one that corresponds to Tatar корал and one that corresponds 
to Tatar чара.  A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in Fig.~\ref{fig:lrx}.

@@

Likewise, the Kazakh word топ can be translated to Tatar as туп \eng{ball} (sometimes
доп in Kazakh), and as төркем \eng{group}.  The bilingual dictionary also has 
the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated 
to Kazakh топ (i.e., analysed), but is never generated.

Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).  The lexical selection rule provided in Fig.~\ref{fig:lrx3} chooses the latter Tatar translation after the adjective рухани \eng{spiritual}.

%Another lexical selection rule is shown in figure~\ref{fig:lrx2}.  This rule ensures that the Kazakh word рет \eng{row, time(s)} is translated as тапкыр \eng{time(s)} when it is used with a number; otherwise it translates as рәт \eng{row}.
%\begin{figure}
%\begin{lstlisting}
%  <rule>
%    <match tags="num"/>
%    <match lemma="рет" tags="n.*">
%        <select lemma="тапкыр"/>
%    </match>
%  </rule>
%\end{lstlisting}
%\caption{A lexical selection rule that selects Tatar тапкыр for Kazakh рет when preceded by a number.}
%\label{fig:lrx2}
%\end{figure}

\begin{figure}
\begin{lstlisting}
  <rule>
    <match lemma="рухани" tags="adj"/>
    <match lemma="өмір" tags="n.*">
        <select lemma="тормыш"/>
    </match>
  </rule>
\end{lstlisting}
\caption{A lexical selection rule that selects тормыш for өмір if preceded by the word рухани.}
\label{fig:lrx3}
\end{figure}

@@

\begin{table}[htbp]
	\centering
	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
			RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
			Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
			\bottomrule
		\end{tabular}
	}\\
	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
		\begin{tabular}{llrrr}
			\toprule
			Corpus          & Tokens & Coverage & stdev\\
			\midrule
			RFERL 2007-2012 & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
			\bottomrule
		\end{tabular}
	}

	\caption{Naïve coverage of the Kazakh-Tatar system}
	\label{table:trimmedcoverage}
\end{table}

@@

For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty (RFERL)'s Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.

For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} a translation of the New Testament, and content from RFERL's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.

@@

%%
%%\begin{table}[htbp]
%%	\centering
%%	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
%%		\begin{tabular}{llrrr}
%%			\toprule
%%			Corpus          & Tokens & Coverage & stdev\\
%%			\midrule
%%			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
%%			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
%%			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
%%			\bottomrule
%%		\end{tabular}
%%	}\\
%%	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
%%		\begin{tabular}{llrrr}
%%			\toprule
%%			Corpus          & Tokens & Coverage & stdev\\
%%			\midrule
%%			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
%%			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
%%			New Testament   & 137K   & 92.54\%  & 1.31\% \\
%%			\bottomrule
%%		\end{tabular}
%%	}
%%	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
%%	\label{table:coverage}
%%\end{table}
\begin{table}
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
  \midrule
  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate over two corpora; OOV is the number of out-of-vocabulary (unknown) words.}
% The development corpus (\texttt{devel}) was used during development 
%       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
%       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.
 %\vspace{-2.5em}
    \label{table:wer-development}
  \end{center}
\end{table}

As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
system\footnote{The coverage of the vanilla transducers is slightly higher.} over 
the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
The coverage over the Wikipedia corpus 
is substantially worse, due to the fact that this corpus is ``dirtier'': it contains orthographical errors, \emph{wiki} code, repetitions, as well as quite a few proper nouns.

@@

We had two small Kazakh corpora along with their postedited translations into Tatar to measure the WER.
The first one (2,457 words total) was a concatenation of an article from RFERL's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, we ran this corpus through the
morphological transducer and manually disambiguated its output.  All the stems in these texts were added to
the system, and all the rules (CG, lexical selection, and transfer) were based on this corpus.
This ``development'' corpus presents an upper bound on the current performance
of the system. %Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN
The testing corpus, used solely for evaluation, was comprised of articles from RFERL's Kazakh-language service, and had a similar size (2,862 words) to the development corpus.
Table~\ref{table:wer-development} presents the WER for both corpora. % http://pastebin.com/7tzGCgMX.

Some of the corrections that were made as part of postediting were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.  A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.  Aside from the irregular behaviour of йөр, which would need its irregularities implemented through special phonotactics in lexc, these issues are all shortcomings of our phonological layer, implemented in twol.

\begin{table}
	\centering
	\begin{tabular}{lll}
		\toprule
		form & current & expected \\\midrule
		ди\tag{<v>}\tag{<tv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & дия & ди \\
		йөр\tag{<v>}\tag{<iv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & йөрә & йөри \\
		кияү\tag{<n>}\tag{<px3sp>}\tag{<nom>} & кияүе & кияве \\
		укы\tag{<v>}\tag{<iv>}\tag{<ger>}\tag{<nom>} & укыу & уку \\\bottomrule
	\end{tabular}
	\caption{Examples of some phonological problems in the Tatar transducer.}
	\label{tab:phonerrors}
\end{table}

@@

@@

%\subsection{Error analysis}

@@

The majority of remaining errors are mostly due to mistakes and gaps in the Tatar morphophonology component, lack of transfer rules to handle some Kazakh compounds, and disambiguation errors.
%\cite{TyersAlperen2010}

@@

To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has near-production-level
coverage, but is rather prototype-level in terms of the number of rules.  Although the impact of this relatively low number of rules on the
quality of translation is extensive (cf., the difference in WER between the development and testing corpora), the outlook is promising and the current results suggest
that a high-quality translation between morphologically-rich agglutinative languages is possible.

We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
increase it to 95\% on the corpora we have, the main work will be improving the quality of translation by adding
more rules, starting with the Constraint Grammar module.
The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.
